---
title: "Multi-Omics Factor Analysis (MOFA2) on Chronic Lymphocytic Leukemia Data"
author: "Based on MOFA2 Tutorial by Ricard Argelaguet and Britta Velten"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: united
    code_folding: show
    highlight: tango
---

```{r setup, include=FALSE}
# ============================================================================
# KNITR SETUP OPTIONS
# ============================================================================
# Configure chunk execution behavior for the entire document
# ============================================================================

knitr::opts_chunk$set(
  echo = TRUE,           # Show code in output
  warning = FALSE,       # Suppress warnings
  message = FALSE,       # Suppress messages
  fig.width = 10,        # Default figure width (inches)
  fig.height = 7,        # Default figure height (inches)
  fig.align = "center",  # Center-align figures
  cache = FALSE          # Don't cache results (ensures reproducibility)
)

# Set random seed for reproducibility across all chunks
set.seed(42)
```

# Introduction to Multi-Omics Factor Analysis (MOFA2)

**MOFA** (Multi-Omics Factor Analysis) is a probabilistic factor analysis model designed for the **unsupervised integration of multi-omic data sets**. In the context of chronic lymphocytic leukemia (CLL), MOFA enables us to identify principal sources of variation across different data modalities (genomic mutations, RNA expression, DNA methylation, and drug response), facilitating:

- **Discovery of latent factors** that represent coordinated variation across multiple omics layers
- **Identification of molecular subtypes** and disease heterogeneity patterns
- **Characterization of clinical markers** such as IGHV mutation status and trisomy 12
- **Interpretation of biological mechanisms** underlying disease progression and drug response

## Biological Context: Chronic Lymphocytic Leukemia

Chronic lymphocytic leukemia (CLL) is a **B-cell malignancy** characterized by significant clinical and molecular heterogeneity. Key molecular markers include:

- **IGHV (Immunoglobulin Heavy-Chain Variable) mutation status**: Mutated IGHV (M-CLL) indicates better prognosis, while unmutated IGHV (U-CLL) is associated with aggressive disease
- **Trisomy 12**: A chromosomal abnormality (extra copy of chromosome 12) associated with intermediate prognosis
- **TP53 mutations and del(17p)**: Associated with poor prognosis and chemoresistance
- **del(13q), del(11q)**: Other common chromosomal abnormalities with varying clinical impacts

The integration of multi-omics data allows us to understand how these genetic lesions translate into phenotypic differences in **DNA methylation patterns, gene expression profiles, and drug sensitivity**.

## Key Approach: Pre-trained Model

This pipeline uses a **pre-trained MOFA model** from the official EBI repository. This approach:

- ✅ Avoids Python/mofapy2 compatibility issues that affect model training
- ✅ Guarantees reproducibility with published results
- ✅ Allows immediate focus on analysis and interpretation
- ✅ Demonstrates the full MOFA2 analysis workflow

---

# 1. Environment Setup and Library Loading

## 1.1 Install Required Packages (Run Once)

```{r install_packages, eval=FALSE}
# ============================================================================
# SECTION 1.1: PACKAGE INSTALLATION
# ============================================================================
# This section installs all required R packages for MOFA2 analysis
# Run this code block only ONCE when setting up your environment
# Set eval=TRUE to run, then set back to eval=FALSE
# ============================================================================

# Install BiocManager if not already installed (required for Bioconductor packages)
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

# Install MOFA2 from Bioconductor
# MOFA2 is the main package for multi-omics factor analysis
BiocManager::install("MOFA2")

# Install MOFAdata package (contains the CLL dataset and pre-trained models)
# ⚠️ Note: This package is large, increase timeout if needed
options(timeout = 600)
BiocManager::install("MOFAdata")

# Install data manipulation and visualization packages
install.packages(c(
  "data.table",    # Fast data manipulation (uses data.table syntax)
  "ggplot2",       # Grammar of graphics visualization
  "tidyverse"      # Collection of data science packages (dplyr, tidyr, etc.)
))
```

## 1.2 Load Required Libraries

```{r load_libraries}
# ============================================================================
# SECTION 1.2: LOAD LIBRARIES
# ============================================================================
# Load all required packages for the analysis
# Each library serves a specific purpose in the workflow
# ============================================================================

# Core MOFA2 packages
library(MOFA2)          # Main package: Multi-Omics Factor Analysis version 2
                        # Provides: create_mofa(), run_mofa(), plot_* functions
library(MOFAdata)       # Data package containing CLL dataset and Reactome gene sets
                        # Provides: CLL_data, CLL_metadata, reactomeGS

# Data manipulation
library(data.table)     # Fast data.frame operations with := and .SD syntax
                        # Used for: efficient data loading and manipulation

# Visualization
library(ggplot2)        # The grammar of graphics for R
                        # Used for: all custom plots and MOFA2 plot modifications

# Tidyverse ecosystem
library(tidyverse)      # Meta-package including:
                        # - dplyr: data manipulation (filter, mutate, summarize)
                        # - tidyr: data reshaping (pivot_longer, pivot_wider)
                        # - purrr: functional programming (map, reduce)
                        # - readr: fast file reading

# Set ggplot2 theme for consistent visualization style
theme_set(theme_bw(base_size = 12))

cat("✓ All libraries loaded successfully!\n")
cat("MOFA2 version:", as.character(packageVersion("MOFA2")), "\n")
```

---

# 2. Data Acquisition and Exploration

## 2.1 Load CLL Multi-Omics Data

The CLL dataset is available through the **MOFAdata** Bioconductor package. It comprises 200 patient samples with four data modalities:

| View | Features | Description |
|------|----------|-------------|
| Somatic Mutations | 69 | Binary indicators of gene mutations (0/1) |
| mRNA Expression | 5,000 | Log-transformed normalized RNA-seq counts |
| DNA Methylation | 4,248 | M-values of most variable CpG sites |
| Drug Response | 310 | Ex vivo drug sensitivity (cell viability) |

```{r load_data}
# ============================================================================
# SECTION 2.1: LOAD CLL MULTI-OMICS DATA
# ============================================================================
# The data is stored as a named list of matrices
# Each matrix has: features in ROWS, samples in COLUMNS
# This is the format expected by create_mofa()
# ============================================================================

# Load the CLL data from MOFAdata package
utils::data("CLL_data")

# Examine the overall structure
cat("=== CLL Data Structure ===\n")
cat("Type:", class(CLL_data), "\n")
cat("Number of omics modalities:", length(CLL_data), "\n")
cat("Modality names:", paste(names(CLL_data), collapse=", "), "\n\n")

# Display dimensions of each modality (features × samples)
cat("=== Data Modality Dimensions (features × samples) ===\n")
for (view in names(CLL_data)) {
  dims <- dim(CLL_data[[view]])
  cat(sprintf("%s: %d features × %d samples\n", view, dims[1], dims[2]))
}
```

## 2.2 Preview Raw Data

```{r preview_data}
# ============================================================================
# SECTION 2.2: PREVIEW RAW DATA
# ============================================================================
# Examine structure and values in each data modality
# Understanding your data is crucial before analysis
# ============================================================================

# Preview mRNA expression data (continuous values, log-transformed)
cat("\n=== Preview of mRNA data (first 5 genes, 5 samples) ===\n")
print(CLL_data$mRNA[1:5, 1:5])

# Preview Mutations data (binary: 0 = wild-type, 1 = mutated)
cat("\n=== Preview of Mutations data (first 5 mutations, 5 samples) ===\n")
print(CLL_data$Mutations[1:5, 1:5])

# Preview Drug response data (continuous: cell viability %)
cat("\n=== Preview of Drug response data (first 5 drugs, 5 samples) ===\n")
print(CLL_data$Drugs[1:5, 1:5])
```

## 2.3 Load Sample Metadata

Clinical metadata provides phenotypic information about patients for biological interpretation of factors.

```{r load_metadata}
# ============================================================================
# SECTION 2.3: LOAD CLINICAL METADATA
# ============================================================================
# Metadata contains clinical information for each patient:
# - Gender: m (male), f (female)
# - age: Patient age in years at sample collection
# - TTT: Time To Treatment (years from sample to next treatment)
# - TTD: Time To Death (years from sample to death)
# - treatedAfter: Whether patient received treatment (TRUE/FALSE)
# - died: Whether patient died during follow-up (TRUE/FALSE)
# - IGHV: IGHV mutation status (0 = unmutated, 1 = mutated)
# - trisomy12: Trisomy 12 status (0 = absent, 1 = present)
# ============================================================================

# Load metadata from the official MOFA repository
CLL_metadata <- tryCatch({
  fread("ftp://ftp.ebi.ac.uk/pub/databases/mofa/cll_vignette/sample_metadata.txt")
}, error = function(e) {
  cat("⚠️ Could not load metadata from FTP. Creating minimal metadata...\n")
  data.frame(sample = colnames(CLL_data$mRNA))
})

# Display metadata summary
cat("=== Metadata Summary ===\n")
cat("Number of samples:", nrow(CLL_metadata), "\n")
cat("Number of variables:", ncol(CLL_metadata), "\n")
cat("Clinical variables:", paste(colnames(CLL_metadata), collapse=", "), "\n\n")

# Display first few rows
cat("=== First 6 Samples ===\n")
print(head(CLL_metadata))

# Check for missing values in key clinical variables
cat("\n=== Missing Value Summary ===\n")
for (col in c("IGHV", "trisomy12", "Gender", "age")) {
  if (col %in% colnames(CLL_metadata)) {
    n_missing <- sum(is.na(CLL_metadata[[col]]))
    cat(sprintf("%s: %d missing (%.1f%%)\n", col, n_missing, 100*n_missing/nrow(CLL_metadata)))
  }
}
```

---

# 3. Load Pre-trained MOFA Model

## 3.1 Why Use a Pre-trained Model?

Training MOFA models requires the **mofapy2** Python package. Due to version incompatibilities between:
- MOFA2 R package (v1.12.1)
- mofapy2 Python package (v0.7.0+)

We use a **pre-trained model** from the official MOFA repository. This model was trained by the MOFA2 developers and guarantees reproducibility with published results.

> 💡 **Learning Point**: In your own analyses, you would train your model using `run_mofa()` after setting up a compatible Python environment.

```{r load_model}
# ============================================================================
# SECTION 3.1: LOAD PRE-TRAINED MOFA MODEL
# ============================================================================
# The pre-trained model is stored as an RDS file on the EBI FTP server
# This model was trained with 15 latent factors on the CLL dataset
# ============================================================================

cat("=== Loading Pre-trained MOFA Model ===\n")
cat("Source: EBI MOFA FTP repository\n")
cat("Model: MOFA2_CLL.rds (trained on 200 CLL samples, 4 views, 15 factors)\n")
cat("Downloading... (this may take a moment)\n\n")

# Load the pre-trained model from the official repository
MOFAobject <- readRDS(url("http://ftp.ebi.ac.uk/pub/databases/mofa/cll_vignette/MOFA2_CLL.rds"))

cat("✓ Model loaded successfully!\n\n")

# Display comprehensive model summary
cat("=== Trained MOFA Model Summary ===\n")
print(MOFAobject)
```

## 3.2 Add Sample Metadata to Model

```{r add_metadata}
# ============================================================================
# SECTION 3.2: ADD SAMPLE METADATA
# ============================================================================
# Attach clinical metadata to the MOFA object for downstream analysis
# This enables coloring/grouping samples by clinical features in plots
# ============================================================================

# Add metadata to the MOFA object
# The samples_metadata<- function requires a data.frame with a 'sample' column
samples_metadata(MOFAobject) <- CLL_metadata

cat("✓ Metadata added to model\n")
cat("Available covariates for analysis:\n")
cat("  ", paste(colnames(samples_metadata(MOFAobject)), collapse=", "), "\n")
```

---

# 4. Model Structure and Interpretation

## 4.1 Understanding MOFA Object Slots

The MOFA object is an S4 class with multiple slots storing data and results.

```{r model_slots}
# ============================================================================
# SECTION 4.1: MOFA OBJECT STRUCTURE
# ============================================================================
# Key slots in the MOFA object:
# - data: Input data matrices (centered at zero)
# - expectations: Posterior expectations of Z (factors) and W (weights)
# - samples_metadata: Clinical information
# - training_stats: Convergence and ELBO values
# ============================================================================

cat("=== MOFA Object Slots ===\n")
cat("The MOFA object contains the following components:\n\n")
slots <- slotNames(MOFAobject)
for (slot in slots) {
  cat("•", slot, "\n")
}

cat("\n=== Most Important Slots for Analysis ===\n")
cat("• data: Input data matrices\n")
cat("• expectations: Factor values (Z) and Feature weights (W)\n")
cat("• samples_metadata: Clinical covariates\n")
cat("• dimensions: Number of views, groups, samples, factors\n")
```

## 4.2 Data Dimensions

```{r data_dims}
# ============================================================================
# SECTION 4.2: DATA DIMENSIONS
# ============================================================================
# Examine the dimensions of input data and learned representations
# ============================================================================

# Data views (input matrices)
cat("=== Input Data Views ===\n")
cat("Views:", paste(names(MOFAobject@data), collapse=", "), "\n\n")

for (view in names(MOFAobject@data)) {
  dims <- dim(MOFAobject@data[[view]]$group1)
  cat(sprintf("• %s: %d features × %d samples\n", view, dims[1], dims[2]))
}
```

## 4.3 Factor and Weight Matrices

```{r factor_weights}
# ============================================================================
# SECTION 4.3: FACTOR AND WEIGHT MATRICES
# ============================================================================
# MOFA learns two key matrices:
# 
# Z (Factors): samples × factors matrix
#   - Each row is a sample
#   - Each column is a latent factor
#   - Values indicate sample position along each factor axis
#   - Interpretation: similar to PCA scores
#
# W (Weights): features × factors matrix (one per view)
#   - Each row is a feature (gene, CpG, drug, mutation)
#   - Each column is a latent factor
#   - Values indicate feature contribution to each factor
#   - Interpretation: similar to PCA loadings
# ============================================================================

# Factor matrix dimensions
cat("=== Factor Matrix (Z) ===\n")
Z_dims <- dim(MOFAobject@expectations$Z$group1)
cat("Dimensions:", Z_dims[1], "samples ×", Z_dims[2], "factors\n")
cat("Interpretation: Each sample has", Z_dims[2], "factor values\n\n")

# Weight matrices dimensions
cat("=== Weight Matrices (W) ===\n")
for (view in names(MOFAobject@expectations$W)) {
  W_dims <- dim(MOFAobject@expectations$W[[view]])
  cat(sprintf("• %s: %d features × %d factors\n", view, W_dims[1], W_dims[2]))
}
cat("\nInterpretation: Each feature has a weight for each factor\n")
cat("High |weight| = strong association with factor\n")
cat("Sign indicates direction of association\n")
```

---

# 5. Variance Decomposition Analysis

## 5.1 Variance Explained per Factor

The most important insight from MOFA is the **variance decomposition plot**, which shows how much variance each factor explains in each data modality.

```{r variance_by_factor, fig.height=8}
# ============================================================================
# SECTION 5.1: VARIANCE EXPLAINED BY FACTOR
# ============================================================================
# This plot is KEY to understanding MOFA results:
# - X-axis: Latent factors (1 to K)
# - Y-axis: Data modalities (views)
# - Color intensity: Percentage of variance explained
#
# INTERPRETATION GUIDE:
# - Factors active in ALL views = shared biological signals
# - Factors active in ONE view = view-specific technical/biological signals
# - High variance = important source of variation
# ============================================================================

cat("=== Variance Decomposition by Factor ===\n")
cat("Interpretation:\n")
cat("• Dark colors = high variance explained (important factor for that view)\n")
cat("• Light colors = low variance explained\n")
cat("• Factors active across multiple views = shared biological signals\n\n")

plot_variance_explained(MOFAobject, max_r2 = 15)
```

### Key Observations:

Based on the variance decomposition:

1. **Factor 1** captures variance across **ALL** modalities → likely represents a major disease driver
2. **Factor 2** is highly specific to the **Drug response** view → drug-specific effect
3. **Factor 3** captures variation across Drugs, mRNA, and Mutations (not Methylation) → another biological signal
4. **Factors 4-15** capture smaller, more specific sources of variation

## 5.2 Total Variance Explained

```{r total_variance}
# ============================================================================
# SECTION 5.2: TOTAL VARIANCE EXPLAINED PER VIEW
# ============================================================================
# This shows how much of the total variance in each view is captured
# by ALL factors combined (K=15 in this model)
#
# Typical values:
# - 10-30% = noisy data or strong non-linearities
# - 30-60% = good fit for most biological data
# - >60% = excellent fit (or possible overfitting)
# ============================================================================

cat("=== Total Variance Explained (all 15 factors) ===\n")
cat("This indicates how well MOFA captures the structure in each view\n\n")

# Extract and plot total variance
plot_variance_explained(MOFAobject, plot_total = TRUE)[[2]]
```

> 💡 **Learning Point**: With only 15 factors, MOFA explains ~54% of Drug response variation and ~42% of mRNA variation. This is excellent for a linear model!

---

# 6. Factor Correlation

```{r factor_correlation}
# ============================================================================
# SECTION 6: FACTOR CORRELATION
# ============================================================================
# A good MOFA model should have largely UNCORRELATED factors
# Each factor should capture a unique source of variation
#
# High correlation between factors suggests:
# - Too many factors (redundancy)
# - Improper data normalization
# - Model convergence issues
# ============================================================================

cat("=== Factor Correlation Matrix ===\n")
cat("Good models have off-diagonal values close to zero\n\n")

plot_factor_cor(MOFAobject)
```

---

# 7. Factor Characterization: Factor 1 (IGHV Mutation)

## 7.1 Association with Clinical Covariates

```{r factor1_association, fig.height=6}
# ============================================================================
# SECTION 7.1: CLINICAL ASSOCIATIONS
# ============================================================================
# Test correlations between factors and clinical covariates
# Helps identify which factors capture clinically relevant variation
# 
# p-values are corrected for multiple testing
# -log10(p-value) is plotted: higher = more significant
# ============================================================================

cat("=== Factor-Covariate Associations ===\n")
cat("Testing: Gender, survival (died), and age\n")
cat("Significance: higher -log10(p-value) = stronger association\n\n")

correlate_factors_with_covariates(MOFAobject, 
  covariates = c("Gender", "died", "age"), 
  plot = "log_pval"
)
```

## 7.2 Plot Factor 1 Values

```{r factor1_values}
# ============================================================================
# SECTION 7.2: FACTOR VALUE DISTRIBUTION
# ============================================================================
# Factor values (Z matrix) represent sample positions along factor axes
# 
# INTERPRETATION:
# - Samples with positive values = high expression of positive-weight features
# - Samples with negative values = high expression of negative-weight features
# - Data is centered → values spread around zero
# ============================================================================

cat("=== Factor 1 Value Distribution ===\n")
cat("Samples are colored by their Factor 1 value\n")
cat("Notice the bimodal distribution suggesting two subgroups\n\n")

plot_factor(MOFAobject, 
  factors = 1, 
  color_by = "Factor1"
)
```

## 7.3 Feature Weights: Mutations View

```{r factor1_mutation_weights}
# ============================================================================
# SECTION 7.3: MUTATION FEATURE WEIGHTS
# ============================================================================
# Feature weights (W matrix) indicate how each feature contributes to a factor
#
# For BINARY mutation data:
# - Positive weight = mutated samples have higher factor values
# - Negative weight = mutated samples have lower factor values
#
# Top-weighted features are the "drivers" of the factor
# ============================================================================

cat("=== Factor 1: Mutation Weights ===\n")
cat("Features with highest |weights| drive the factor\n")
cat("IGHV mutation should appear as a top feature\n\n")

plot_weights(MOFAobject,
  view = "Mutations",
  factor = 1,
  nfeatures = 10,    # Show top 10 features
  abs = FALSE        # Show signed weights (not absolute values)
)
```

> 🎯 **Key Finding**: IGHV mutation has the **largest weight** for Factor 1! This confirms Factor 1 captures IGHV-related biology.

## 7.4 Visualize IGHV Status

```{r factor1_ighv}
# ============================================================================
# SECTION 7.4: FACTOR 1 BY IGHV STATUS
# ============================================================================
# Validate that Factor 1 separates samples by IGHV mutation status
# This confirms the biological interpretation from the weight analysis
# ============================================================================

cat("=== Factor 1 Values by IGHV Mutation Status ===\n")
cat("IGHV = 0: Unmutated (aggressive, poor prognosis)\n")
cat("IGHV = 1: Mutated (indolent, better prognosis)\n\n")

plot_factor(MOFAobject, 
  factors = 1, 
  color_by = "IGHV",
  dodge = TRUE,      # Separate groups horizontally
  add_violin = TRUE  # Show distribution shape
)
```

## 7.5 mRNA Expression Weights

```{r factor1_mrna_weights, fig.height=8}
# ============================================================================
# SECTION 7.5: mRNA FEATURE WEIGHTS
# ============================================================================
# Examine which genes are associated with Factor 1
# 
# Genes with positive weights: higher expression in IGHV-mutated samples
# Genes with negative weights: higher expression in IGHV-unmutated samples
#
# These are potential biomarkers and downstream targets
# ============================================================================

cat("=== Factor 1: Top mRNA Weights ===\n")
cat("These genes show the strongest association with IGHV status\n\n")

plot_weights(MOFAobject,
  view = "mRNA",
  factor = 1,
  nfeatures = 15,
  abs = FALSE
)
```

## 7.6 Gene Expression Heatmap

```{r factor1_heatmap}
# ============================================================================
# SECTION 7.6: EXPRESSION HEATMAP
# ============================================================================
# Visualize expression patterns of top Factor 1 genes across samples
# 
# denoise = TRUE: Shows reconstructed data (noise removed by model)
# scale = "row": Z-score normalization per gene for visualization
# ============================================================================

cat("=== Factor 1: Gene Expression Heatmap ===\n")
cat("Showing top 25 genes, samples ordered by Factor 1 value\n")
cat("denoise=TRUE removes noise for clearer patterns\n\n")

plot_data_heatmap(MOFAobject, 
  view = "mRNA",
  factor = 1,  
  features = 25,         # Top 25 genes by weight
  denoise = TRUE,        # Show denoised (reconstructed) data
  cluster_rows = FALSE,  # Keep genes in weight order
  cluster_cols = FALSE,  # Keep samples in factor order
  show_rownames = TRUE,
  show_colnames = FALSE,
  scale = "row"          # Z-score normalization per gene
)
```

---

# 8. Factor Characterization: Factor 3 (Trisomy 12)

## 8.1 Mutation Weights

```{r factor3_weights}
# ============================================================================
# SECTION 8.1: FACTOR 3 MUTATION WEIGHTS
# ============================================================================
# Identify which genetic lesions drive Factor 3
# ============================================================================

cat("=== Factor 3: Mutation Weights ===\n")
cat("Looking for genetic drivers of Factor 3\n\n")

plot_weights(MOFAobject, 
  view = "Mutations", 
  factor = 3, 
  nfeatures = 10,
  abs = FALSE
)
```

> 🎯 **Key Finding**: Trisomy 12 is the **major driver** of Factor 3! This is the second most important clinical marker in CLL.

## 8.2 Factor Values by Trisomy 12

```{r factor3_trisomy12}
# ============================================================================
# SECTION 8.2: FACTOR 3 BY TRISOMY 12 STATUS
# ============================================================================
# Validate Factor 3 separates samples by trisomy 12 status
# ============================================================================

cat("=== Factor 3 Values by Trisomy 12 Status ===\n")
cat("Trisomy 12 = 0: Normal chromosome 12\n")
cat("Trisomy 12 = 1: Extra copy of chromosome 12\n\n")

plot_factor(MOFAobject, 
  factors = 3, 
  color_by = "trisomy12",
  dodge = TRUE,
  add_violin = TRUE
)
```

## 8.3 Drug Response Signatures

```{r factor3_drugs}
# ============================================================================
# SECTION 8.3: DRUG RESPONSE PATTERNS
# ============================================================================
# Examine how drug sensitivity relates to Factor 3 (Trisomy 12)
# This reveals potential therapeutic vulnerabilities
# ============================================================================

cat("=== Factor 3: Drug Response Patterns ===\n")
cat("Showing drugs with highest positive weights\n")
cat("Potential therapeutic targets for Trisomy 12 patients\n\n")

plot_data_scatter(MOFAobject, 
  view = "Drugs",
  factor = 3,  
  features = 4,           # Top 4 drugs
  sign = "positive",      # Positive weight drugs
  color_by = "trisomy12"
) + labs(y = "Drug response (cell viability)")
```

---

# 9. Patient Stratification: Combined Factor Analysis

## 9.1 Factor 1 vs Factor 3 Scatter Plot

This is one of the **most important visualizations** in MOFA analysis. It stratifies patients into molecular subgroups based on their multi-omic profiles.

```{r combined_factors, fig.height=8}
# ============================================================================
# SECTION 9.1: PATIENT STRATIFICATION
# ============================================================================
# Plot Factor 1 (IGHV) vs Factor 3 (Trisomy 12) to classify patients
# 
# This creates 4 molecular subgroups:
# 1. IGHV-mutated, no Trisomy 12 (best prognosis)
# 2. IGHV-mutated, Trisomy 12
# 3. IGHV-unmutated, no Trisomy 12
# 4. IGHV-unmutated, Trisomy 12 (worst prognosis)
#
# These subgroups predict drug response and survival!
# ============================================================================

cat("=== Patient Stratification by MOFA Factors ===\n")
cat("Color: IGHV mutation status\n")
cat("Shape: Trisomy 12 status\n")
cat("Dashed lines separate the 4 molecular subgroups\n\n")

p <- plot_factors(MOFAobject, 
  factors = c(1, 3), 
  color_by = "IGHV",
  shape_by = "trisomy12",
  dot_size = 2.5,
  show_missing = TRUE
)

# Add reference lines to define quadrants
p <- p + 
  geom_hline(yintercept = -1, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = -0.5, linetype = "dashed", color = "gray50") +
  labs(
    title = "CLL Patient Stratification by MOFA Factors",
    subtitle = "Based on IGHV mutation status and Trisomy 12",
    x = "Factor 1 (IGHV mutation axis)",
    y = "Factor 3 (Trisomy 12 axis)"
  )

print(p)
```

### Clinical Significance

This plot demonstrates the power of multi-omics integration:

1. **Factor 1** (x-axis) separates patients by **IGHV mutation status**
2. **Factor 3** (y-axis) separates patients by **Trisomy 12 status**
3. The four quadrants represent **clinically distinct subgroups** with different:
   - Disease progression rates
   - Drug sensitivity profiles
   - Survival outcomes

> 🏥 **Clinical Implication**: These molecular subgroups can guide **precision medicine** treatment decisions!

---

# 10. Gene Set Enrichment Analysis (GSEA)

## 10.1 Load Reactome Gene Sets

```{r gsea_setup}
# ============================================================================
# SECTION 10.1: GENE SET SETUP
# ============================================================================
# Load pre-defined gene sets from Reactome database
# These represent biological pathways and processes
# 
# The gene set matrix is BINARY:
# - Rows = pathways
# - Columns = genes
# - 1 = gene belongs to pathway, 0 = gene not in pathway
# ============================================================================

cat("=== Loading Reactome Gene Sets ===\n")

# Load Reactome pathways from MOFAdata
utils::data(reactomeGS)

cat("Number of pathways:", nrow(reactomeGS), "\n")
cat("Number of genes:", ncol(reactomeGS), "\n\n")

cat("Example pathway names:\n")
print(head(rownames(reactomeGS)))
```

## 10.2 Run Enrichment Analysis

```{r gsea_run}
# ============================================================================
# SECTION 10.2: RUN GSEA
# ============================================================================
# Test which pathways are enriched in each factor's top genes
# 
# Parameters:
# - feature.sets: Binary gene set matrix
# - view: Which MOFA view to analyze (must be mRNA for pathway analysis)
# - factors: Which factors to test
# - set.statistic: How to summarize gene weights into pathway scores
# - statistical.test: How to test significance
# ============================================================================

cat("=== Running Gene Set Enrichment Analysis ===\n")
cat("Testing Reactome pathways against Factor weights\n")
cat("This may take a moment...\n\n")

enrichment_results <- run_enrichment(MOFAobject, 
  feature.sets = reactomeGS, 
  view = "mRNA",
  factors = 1:5,                    # Test first 5 factors
  set.statistic = "mean.diff",      # Use mean weight difference
  statistical.test = "parametric"   # Parametric t-test
)

cat("✓ Enrichment analysis complete!\n")
```

## 10.3 Plot Enrichment Results

```{r gsea_plot, fig.height=10}
# ============================================================================
# SECTION 10.3: VISUALIZE GSEA RESULTS
# ============================================================================
# Plot top enriched pathways for Factor 1
# ============================================================================

cat("=== Factor 1: Top Enriched Pathways ===\n")

tryCatch({
  plot_enrichment(enrichment_results, 
    factor = 1, 
    max.pathways = 15,
    alpha = 0.1
  )
}, error = function(e) {
  cat("Note: No pathways met significance threshold (alpha=0.1)\n")
  cat("Showing top pathways by adjusted p-value:\n\n")
  
  # Manual extraction of results
  if (!is.null(enrichment_results$pval.adj)) {
    results_df <- data.frame(
      pathway = rownames(enrichment_results$pval.adj),
      pval_adj = enrichment_results$pval.adj[, 1]
    )
    results_df <- results_df[order(results_df$pval_adj), ]
    print(head(results_df, 15))
  }
})
```

---

# 11. Dimensionality Reduction (UMAP)

```{r umap, fig.height=7}
# ============================================================================
# SECTION 11: UMAP VISUALIZATION
# ============================================================================
# UMAP (Uniform Manifold Approximation and Projection) reduces the
# 15-dimensional factor space to 2D for visualization
#
# Samples that cluster together have similar multi-omic profiles
# Color/shape by clinical features to identify subgroups
# ============================================================================

cat("=== UMAP of MOFA Factors ===\n")
cat("Reducing 15 factors to 2D for visualization\n\n")

# Run UMAP on factor values
set.seed(42)
MOFAobject <- run_umap(MOFAobject)

# Plot UMAP colored by clinical features
plot_dimred(MOFAobject, 
  method = "UMAP",
  color_by = "IGHV",
  shape_by = "trisomy12",
  dot_size = 3
) + 
  labs(
    title = "UMAP Projection of MOFA Factors",
    subtitle = "Samples colored by IGHV status, shaped by Trisomy 12"
  )
```

---

# 12. Session Information

```{r session_info}
# ============================================================================
# SECTION 12: SESSION INFORMATION
# ============================================================================
# Document R version and package versions for reproducibility
# ============================================================================

cat("=== Session Information ===\n")
cat("For reproducibility, here are the package versions used:\n\n")

sessionInfo()
```

---

# Summary and Key Takeaways

## What We Learned

This analysis demonstrated how **MOFA+ integrates multiple omics** to discover biologically meaningful patterns:

### Key Findings

| Factor | Driver | Clinical Relevance |
|--------|--------|-------------------|
| Factor 1 | IGHV mutation | Major prognostic marker; separates M-CLL from U-CLL |
| Factor 3 | Trisomy 12 | Second prognostic marker; affects drug sensitivity |

### Technical Insights

1. **Variance Decomposition**: Revealed which factors capture shared vs view-specific variation
2. **Feature Weights**: Identified molecular drivers (genes, mutations) underlying each factor
3. **Patient Stratification**: Classified 200 CLL patients into 4 molecular subgroups
4. **GSEA**: Linked factors to biological pathways

### Clinical Implications

MOFA factors can be used for:
- ✅ **Precision medicine**: Stratify patients for targeted therapy
- ✅ **Biomarker discovery**: Identify genes predictive of drug response
- ✅ **Survival prediction**: Factors correlate with time-to-treatment
- ✅ **Drug repurposing**: Find drugs effective in specific molecular subgroups

## Next Steps for Your Own Analysis

1. **Prepare your data**: Format as list of matrices (features × samples)
2. **Set up Python**: Create compatible environment with mofapy2==0.6.x
3. **Train model**: Use `run_mofa()` with your data
4. **Interpret factors**: Follow the workflow demonstrated here

---

*Analysis completed on `r Sys.Date()`*
